---
title: "causality_demo"
author: "Klaus Frieler"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("dag_exercise.R")
global_seed <- 786543 
simple <- simulate_simple_example()
```

## A contrived example

We simulated a data set based on the following DAG.
```{r simple_data_step_dag}
ggdag(simple_dag_step1, layout = "time_ordered") +
  theme_dag()   +
  labs(title = "Contrived Example")


```

We simulated an observational experiment with a treatment t (0, 1, e.g,  "playing music last week") and an continuous outcome y (e.g., IQ). We have some data for it. What is the Average Treatment Effect? 

```{r simple_data_step1}

simple %>%  select(t, y) %>% psych::pairs.panels() 
#simple %>% with(., table(t))
simple %>% boxplot( y ~ t, data = .)
simple %>% group_by(t) %>% summarise(m = mean(y))
simple %>% 
  rstatix::t_test( y ~ t) %>% select(-c(group1, group2)) %>% 
  bind_cols(simple %>% 
  rstatix::cohens_d( y ~ t) %>% select(effsize) 
)

```
Great, a really strong effect. But we know and measured two covariates that might influence the outcome as well (e.g., SES, Working Memory)

```{r simple_data_step2}
ggdag(simple_dag_step2, layout = "time_ordered") +
  theme_dag()   +
  labs(title = "Contrived Example")


```

```{r simple_data_reg_cov}

simple %>% lm(y ~ t + l1 + l2, data = .) %>% summary()
  

```

So, this was successful, the effect size was halfed with ATE = 9.65. But actually, we might tend to think, that the treatment is also affected by the covariates. Can we predict "treatment" from covariates?

```{r treatment_model}

ps_mod <- simple %>% glm(t ~ l1 + l2, family = "binomial", data = .) 
sjPlot::tab_model(ps_mod)
ps_mod %>% summary()
ps_mod %>% performance::r2_tjur()

```
This seems to be the case. Let's try some propensity score weighting.

```{r prop_scores}
ps <- ps_mod %>% predict(type = "response")
hist(ps)
as.integer(ps > .5) %>% table()
table(simple$t)
```

Let's try IPW  regression using propensity score, without covariates.

```{r simple_data_reg_prop}

simple_ext <- simple %>% group_by(t) %>% 
  mutate(rel_f = n()/nrow(.)) %>%
  ungroup() %>% 
  mutate(weights = t * ps + (1 -t) * (1 - ps), 
         stab_weights = rel_f * (t * ps + (1 -t) * (1 - ps)))
         
simple_ext$weights %>% hist()
ipw_prop1 <- simple_ext %>% 
           lm(y ~ t, weights = weights, data = .) %>% summary()
ipw_prop1  
ipw_prop2 <- simple_ext %>% 
           lm(y ~ t, weights = stab_weights, data = .) %>% summary()
ipw_prop2  

```

So, this didn't work. Let's add covariates

```{r simple_data_reg_prop_cov}

ipw_prop3 <- simple_ext %>% 
           lm(y ~ t + l1 + l2, weights = weights, data = .) %>% summary()
ipw_prop3
```
Not so much changes.

Let's try matching, "full" and then CEM.

```{r simple_data_reg_match}
  match_full <- simple %>% 
    MatchIt::matchit(t ~ l1 + l2, method = "full", 
                     distance = "glm", 
                     link = "logit", data  =.) 

  match_cem <- simple %>% 
    MatchIt::matchit(t ~ l1 + l2, 
                     method = "cem", data  =.)
  simple_full <- match_data(match_full)
  simple_cem <- match_data(match_cem)
  fit_full <- simple_full %>% 
    lm(y ~ t + l1 + l2, 
       weights = weights, 
       data = .)
  
  fit_full0 <- simple_full %>% 
    lm(y ~ t, 
       weights = weights, 
       data = .)
  fit_cem <- simple_cem %>% 
    lm(y ~ t + l1 + l2, 
       weights = weights, 
       data = .)
  
  fit_cem0 <- simple_cem %>% 
    lm(y ~ t, 
       weights = weights, 
       data = .)
sjPlot::plot_models(fit_cem, fit_cem0, fit_full, fit_full0, lm(y ~t, data = simple), m.labels = c("CEM", "CEM0", "FULL", "FULL0", "NAIVE"))
```
Finally, let's try TMLE.

```{r simple_data_reg_tmle}
mod1 <- simple %>%
  lmtp_tmle(
    trt = "t",
    outcome = "y",
    outcome_type = "continuous",
    baseline = c("l1", "l2"),
    shift = static_binary_on
  )
mod0 <- simple %>% lmtp_tmle(
  trt = "t",
  outcome = "y",
  outcome_type = "continuous",
  baseline = c("l1", "l2"),
  shift = static_binary_off
)
lmtp_contrast(mod1, ref = mod0)

```

So what is the truth? Here is the code to generate the simulated data. The true ATE is 10.

<pre><code>
simulate_simple_example <- function(n = 1000, 
                                    seed = global_seed, 
                                    threshold = .1, 
                                    cov_rel = 0){
  set.seed(seed)
  simdat <- tibble(l1 = rnorm(n, 10, 5))
  
  #use some normal distribution to defined treatment
  a.lin <- simdat$l1 - 10
  b.lin <- rnorm(n, 2, 5)
  
  #c.lin will be unobserved
  c.lin <- rnorm(n, 1, 5)
  
  #logistic experssion to defone probality to be treated 
  pa <- exp(a.lin + b.lin + c.lin)/(1 + exp(a.lin + b.lin + c.lin))
  
  #simulate binary treatment based on probability to treat, which in turn is based on 2 obversed and on unobserved covariate
  simdat$t <- rbinom(n, 1, prob = pa)
  
  #add reandom treatment for whatever reasons
  simdat$r <- sample(c(-1,1), nrow(simdat), replace = T)
  
  #outcome
  simdat$y <- 10 * simdat$t + 2 * simdat$l1 + rnorm(n, -10, 5)
  
  #Covariate l1 with measurement error based on cov_rel (covariate reliability)
  simdat$l1_me <- simdat$l1 + rnorm(n, 0, cov_rel) 
  
  #covariate l2
  simdat$l2 <- b.lin  

  #Covariate l2 with measurement error, based on cov_rel)
  simdat$l2_me <- b.lin + rnorm(n, 0, cov_rel) 
  
  #change threshold/2 percents of treatments  
  shuffle_idx <- which(runif(nrow(simdat)) > (1 - threshold))
  simdat$t_me <- simdat$t
  simdat$t_me[shuffle_idx] <- sample(c(0, 1), length(shuffle_idx), replace = T)
  
  #add a collider, for education purposes
  simdat <- simdat %>% mutate(z = scale((y + l1 * l2))[,1] + rnorm(nrow(simdat), 10, 1))
  #add ids for completeness
  simdat$id <- 1:nrow(simdat)
  simdat
}

</code></pre>
`

This includes also some more variables, i.e., measurement error for treatment `t_me`, for covaviates `l1_me`, `l2_me' as well as a collider `z` depending on `y`, `l1` and `l2`.

Let's try different estimations methods for different combinations of theses variables and compare. We used clean (trt = `clean`) or noisy treatment (about `5%` wrong treatments, trt = `noisy`), with or without measurement error for `l1` and `l2` (with_measurement error `none`or `some`), and the following methods 

  * simple linear regression (method = `none`)
  * tmle (method = `tmle`)
  * full matching fromt the `matchIt` package (method = `fuil matching`) with or without weights (weights = `none` or weights = `IPW`)
  * Coarsend Exact Matching (CEM) from the `matchIt` package (method = `CEM`) with or without weights (weights = `none` or weights = `IPW`)

with the following model variants

  * covariates `l1`and `l2` (spec = `covariates`)
  * wrong covariates `z` and `l2` (spec = `wrong covariates`)
  * covariates `l1`, `l2`and collider `z` (spec = `covariates + collider`)

`ref` is the mean in the non-treated group, and `shift` is the mean in the treated group, either raw means or as estimated by the `lmtp_tmle` function. `ATE`is the estimated average treatment effect.

```{r simple_data_ATE_comp}
comp_ATE_raw <- comp_ATE_simple_dags(with_measurement_error = F, data = simple)
comp_ATE_me <- comp_ATE_simple_dags(with_measurement_error = T, data = simple)
comp_ATE <- bind_rows(comp_ATE_raw, comp_ATE_me) %>% 
  rename(ATE = estimate) %>% mutate(difference = ATE - 10)  %>% 
  select(ATE, difference, ref, shift, trt, measurement_error = with_measurement_error, 
         method, spec, trt, weights, everything())  

comp_ATE %>%   arrange(abs(difference)) %>% sjPlot::tab_df()
``` 
