---
title: "causality_demo"
author: "Klaus Frieler"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("dag_exercise.R")
simple <- simulate_simple_example()
```

## A contrived example

We simulated a data set based on the following DAG.
```{r simple_data_step1}
ggdag(simple_dag_step1, layout = "time_ordered") +
  theme_dag()   +
  labs(title = "Contrived Example")


```

We simulated an observational experiment with a treatment t (0, 1, e.g,  "playing music last week") and an continuous outcome y (e.g., IQ). We have some data for it. What is the Average Treatment Effect? 

```{r simple_data_step1}

simple %>%  select(t, y) %>% psych::pairs.panels() 
#simple %>% with(., table(t))
simple %>% boxplot( y ~ t, data = .)
simple %>% group_by(t) %>% summarise(m = mean(y))
simple %>% 
  rstatix::t_test( y ~ t) %>% select(-c(group1, group2)) %>% 
  bind_cols(simple %>% 
  rstatix::cohens_d( y ~ t) %>% select(effsize) 
)

```
Great, a really strong effect. But we know and measured two covariates that might influence the outcome as well (e.g., SES, Working Memory)

```{r simple_data_step2}
ggdag(simple_dag_step2, layout = "time_ordered") +
  theme_dag()   +
  labs(title = "Contrived Example")


```

```{r simple_data_reg_cov}

simple %>% lm(y ~ t + l1 + l2, data = .) %>% summary()
  

```

So, this was successful, the effect size was halfed with ATE = 9.65. But actually, we might tend to think, that the treatment is also affected by the covariates. Can we predict "treatment" from covariates?

```{r treatment_model}

ps_mod <- simple %>% glm(t ~ l1 + l2, family = "binomial", data = .) 
sjPlot::tab_model(ps_mod)
ps_mod %>% summary()
ps_mod %>% performance::r2_tjur()

```
This seems to be the case. Let's try some propensity score weighting.

```{r prop_scores}
ps <- ps_mod %>% predict(type = "response")
hist(ps)
as.integer(ps > .5) %>% table()
table(simple$t)
```

Let's try IPW  regression using propensity score, without covariates.

```{r simple_data_reg_prop}

simple_ext <- simple %>% group_by(t) %>% 
  mutate(rel_f = n()/nrow(.)) %>%
  ungroup() %>% 
  mutate(weights = t * ps + (1 -t) * (1 - ps), 
         stab_weights = rel_f * (t * ps + (1 -t) * (1 - ps)))
         
simple_ext$weights %>% hist()
ipw_prop1 <- simple_ext %>% 
           lm(y ~ t, weights = weights, data = .) %>% summary()
ipw_prop1  
ipw_prop2 <- simple_ext %>% 
           lm(y ~ t, weights = stab_weights, data = .) %>% summary()
ipw_prop2  

```

So, this didn't work. Let's add covariates

```{r simple_data_reg_prop}

ipw_prop3 <- simple_ext %>% 
           lm(y ~ t + l1 + l2, weights = weights, data = .) %>% summary()
ipw_prop3
```
Not so much changes.

Let's try matching, "full" and then CEM.

```{r simple_data_reg_match}
  match_full <- simple %>% 
    MatchIt::matchit(t ~ l1 + l2, method = "full", 
                     distance = "glm", 
                     link = "logit", data  =.) 

  match_cem <- simple %>% 
    MatchIt::matchit(t ~ l1 + l2, 
                     method = "cem", data  =.)
  simple_full <- match_data(match_full)
  simple_cem <- match_data(match_cem)
  fit_full <- simple_full %>% 
    lm(y ~ t + l1 + l2, 
       weights = weights, 
       data = .)
  
  fit_full0 <- simple_full %>% 
    lm(y ~ t, 
       weights = weights, 
       data = .)
  fit_cem <- simple_cem %>% 
    lm(y ~ t + l1 + l2, 
       weights = weights, 
       data = .)
  
  fit_cem0 <- simple_cem %>% 
    lm(y ~ t, 
       weights = weights, 
       data = .)
sjPlot::plot_models(fit_cem, fit_cem0, fit_full, fit_full0, lm(y ~t, data = simple), m.labels = c("CEM", "CEM0", "FULL", "FULL0", "NAIVE"))
```
Finally, let's try TMLE.

```{r simple_data_reg_tmle}
mod1 <- simple %>%
  lmtp_tmle(
    trt = "t",
    outcome = "y",
    outcome_type = "continuous",
    baseline = c("l1", "l2"),
    shift = static_binary_on
  )
mod0 <- simple %>% lmtp_tmle(
  trt = "t",
  outcome = "y",
  outcome_type = "continuous",
  baseline = c("l1", "l2"),
  shift = static_binary_off
)
lmtp_contrast(mod1, ref = mod0)

```

So what is the truth?

```{r simple_data_reg_tmle}
simulate_simple_example <- function(n = 1000, 
                                    seed = global_seed, 
                                    threshold = .1, 
                                    cov_rel = 0){
  set.seed(global_seed)
  simdat <- tibble(l1 = rnorm(n, 10, 5))
  a.lin <- simdat$l1 - 10
  b.lin <- rnorm(n, 2, 5)
  c.lin <- rnorm(n, 1, 5)
  pa <- exp(a.lin + b.lin + c.lin)/(1 + exp(a.lin + b.lin + c.lin))
  simdat$a <- rbinom(n, 1, prob = pa)
  simdat$r <- sample(c(-1,1), nrow(simdat), replace = T)
  simdat$y <- 10 * simdat$a + 2 * simdat$l1 + rnorm(n, -10, 5)
  simdat$l1_me <- simdat$l1 + rnorm(n, 0, cov_rel) 
  simdat$l2 <- b.lin  
  simdat$l2_me <- b.lin + rnorm(n, 0, cov_rel) 
  shuffle_idx <- which(runif(nrow(simdat)) > (1 - threshold))
  simdat$t <- simdat$a
  simdat$a[shuffle_idx] <- sample(c(0, 1), length(shuffle_idx), replace = T)
  #simdat$a[abs(b.lin-20) > ] <- sample(c(0, 1), length(shuffle_idx), replace = T)
  simdat$id <- 1:nrow(simdat)
  simdat %>% mutate(z = scale((y + l1 * l2))[,1]+ rnorm(nrow(simdat), 10, 1))
}
```

